SAT Solving for Vesicle Traffic Systems in Cells

### 2017-07-20 paper/bio_MTedit.tex
sad = shiromani_akali_dal

\begin{enumerate}
\item We determine which molecules are active on every compartment or vesicle.
\item For every vesicle fusing to a compartment, we determine whether there exists an active pair (one molecule on the vesicle, one on the compartment) which drives that fusion event.
\item For every vesicle-compartment pair where the vesicle does not fuse to the compartment, we verify that there is no pairing of active molecules on the
vesicle and compartment that could drive their fusion.
\item We verify that every molecular type entering a compartment also leaves the compartment, and also that every molecular type entering a set of compartments also leaves that set; this is the steady state condition.
\end{enumerate}

@@

\begin{example}
%
In figure~\ref{fig:M1}, we present a VTS that has 3 compartments and 8 molecules, and a corresponding pairing matrix. Molecules are numbered 0-7. In the VTS, labels are a string of molecule ids, and an overline over a an id indicates that the molecule is active. This VTS is an example of one where every molecule on the node is active, and the activity of the molecules on the edges are controlled by presence of the other molecules. On the right side of the figure, we show the pairing matrix. An entry 1 represents pairing between molecules; $\times$ represents no pairing. Rows are labels on vesicles, columns are labels on compartments. Every molecule flows on a cycle, ensuring steady state. This is a 3-connected graph.
\end{example}

\subsection{Modeling and Symbolic Analysis of VTS: An Overview}
\label{subsec:graphmodel}
Since a VTS are a transport graph, it is but natural to formally model VTS as graphs (as used in computer science) with their nodes denoting compartments and labeled edges denoting transport vesicles with labels denoting the set of molecules being transported.  The budding and fusion regulations can be represented as matching tables over sets of molecules, i.e., formally as a boolean function that given requisite labels of nodes and vesicles that returns true if the required regulations are met.  Given such a graph model of a VTS, then properties, such as $k$-connectedness or steady state condition, can be formally defined as constraints, i.e., first-order logical formulas, over graphs and function types.  Note that these formulas define among other things constraints over paths between nodes in the graph model of VTS.  For example, the steady state condition, described informally earlier, can be defined by the formula shown in Listing~1.1.

\begin{lstlisting}[mathescape,
 breaklines,
  rulecolor=\color{black},
  frame=single,
  caption= \textbf{Steady state specification.},
  moredelim={[is][keywordstyle]{@@}{@@}}
]
# Every edge E is represented by a label; labelE (a bitvector).
# E(x,y): Edge between node x and y. E.source = x, E.target = y    
# labelE(m): mth molecule is present on edge E.
# Edge(m)(x,y): Edge between node x and y with labelE(m).                 
# seq(a1..an): each ak represents a node. len(seq) is |seq|.
# N: total number of nodes, = : Mathematical equality.

@@$\forall$ z $\in$ {Nodes} $\forall$ E $\in$ {Edges}: E.source = z
    $\forall$ m $\in$ {Molecules}: labelE(m) $\supset$
       ($\exists$ a seq(a1..an): (2 $\leq$ |seq| $\leq$ N)  $\wedge$
            ((a1 = z) $\wedge$ (a2 = E.target) $\wedge$
             (($\forall$ k from 1 to |seq| - 1) $\supset$ Edge(m)(ak,ak+1)) $\wedge$
             (Edge(m)(an,a1))) @@

\end{lstlisting}    
%\begin{itemize}
%\item \srivas{Detailed BIO to Graph problem}
%\end{itemize}

\begin{lstlisting}[mathescape,
  breaklines,
  rulecolor=\color{black},
  frame=single, 
  caption= \textbf{Maximal Connectedness (LGC) Property.},
  moredelim={[is][keywordstyle]{@@}{@@}}
] 
# C1: Steady state condition.
# C2, C3: Fusion rules.
# C4: Graph is 4-connected.
# G: Graph, f: Constrained function.
# $\supset$: material conditional (if..then), $\wedge$: Logical and.
$\forall$: Universal quantifier, $\exists$: Existential quantifier.
# For every 4-connected graph following conditions must be valid.  
@@VALID ($\forall$ G: C4(G) $\supset$  ($\exists$ f: C1(G) $\wedge$ C2(G,f) $\wedge$ C3(G,f) )) @@

\end{lstlisting}

%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

>>>>

\begin{enumerate}
\item We determine which molecules are active on every compartment or vesicle.
\item For every vesicle fusing to a compartment, we determine whether there exists an active pair (one molecule on the vesicle, one on the compartment) which drives that fusion event.
\item For every vesicle-compartment pair where the vesicle does not fuse to the compartment, we verify that there is no pairing of active molecules on the
vesicle and compartment that could drive their fusion.
\item We verify that every molecular type entering a compartment also leaves the compartment, and also that every molecular type entering a set of compartments also leaves that set; this is the \textbf{stability condition}.
%steady state condition.
\end{enumerate}

%\subsection{Modeling and Symbolic Analysis of VTS: An Overview}
\label{subsec:graphmodel}
Since a VTS are a transport graph, it is but natural to formally model VTS as graphs (as used in computer science) with their nodes denoting compartments and labeled edges denoting transport vesicles with labels denoting the set of molecules being transported.  The budding and fusion regulations can be represented as matching tables over sets of molecules, i.e., formally as a boolean function that given requisite labels of nodes and vesicles that returns true if the required regulations are met.  Given such a graph model of a VTS, then properties, such as $k$-connectedness or steady state condition, can be formally defined as constraints, i.e., first-order logical formulas, over graphs and function types.  Note that these formulas define among other things constraints over paths between nodes in the graph model of VTS. 
% For example, the steady state condition, described informally earlier, can be defined by the formula shown in Listing~1.1.

%\begin{itemize}
%\item \srivas{Detailed BIO to Graph problem}
%\end{itemize}

@@

\begin{example}
%
In figure~\ref{fig:M1}, we present a VTS that has 3 compartments and 8 molecules, and a corresponding pairing matrix. Molecules are numbered 0-7. In the VTS, labels are a string of molecule ids, and an overline over a an id indicates that the molecule is active. This VTS is an example of one where every molecule on the node is active, and the activity of the molecules on the edges are controlled by presence of the other molecules. On the right side of the figure, we show the pairing matrix. An entry 1 represents pairing between molecules; $\times$ represents no pairing. Rows are labels on vesicles, columns are labels on compartments. Every molecule flows on a cycle, ensuring steady state. This is a 3-connected graph.
\end{example}
%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

### 2017-07-20 paper/intro_MTedit.tex
sad = shiromani_akali_dal

\begin{enumerate}
\item The encoding used to model VTSs was not fine-tuned to the structure of the problem.
\item The queries to be solved involved quantified formulas, i.e., we needed to solve queries of the form ``\emph{for all} graphs does there \emph{exist} SNARE regulation constraints that satisfy cellular steady state conditions.
\end{enumerate}

As CBMC does not directly support solving such quantified formulas, we had to use a technique that combined non-determinism and partial enumeration that is not as efficient.\newline

In this paper, we have developed a novel SAT encoding for searching a VTS that satisfy the given properties. We improved the encoding of some of the conditions for VTSs. In the earlier implementation, reachability was encoded by enumeration of paths. In our current encoding, we have recursively defined the reachability condition such that that we avoid the exponential enumeration. We have also encoded $k$-connectivity in a similar fashion and avoided a blowup in the formula size. Furthermore, our encoding supports several variations of the conditions. We have implemented the encoding using Z3~\cite{z3} python API and searched for VTS satisfying various properties upto size 14-18 nodes as compare earlier tool that could only search graph upto size 8 nodes. Our tool is useful to cell biologists since VTSs of real cells have approximately ten compartments.  \\

>>>>

\begin{enumerate}
\item The encoding used to model VTSs was not fine-tuned to the structure of the problem.
\item The \textbf{specification involved enumeration of all the possible paths in order to satisfy both the graph connectivity and the stability condition (refer section 3).}
%quantified formulas, i.e., we needed to solve queries of the form ``\emph{for all} graphs does there \emph{exist} SNARE regulation constraints that satisfy cellular steady state conditions.
\end{enumerate}

The problem of encoding graph connectivity and stability condition in CBMC was solved combined non-determinism and partial enumeration that is not as efficient.\newline

\todo{REQUIRE A REWRITE FLOW PROBS}
In this paper, we have developed a novel SAT encoding for searching a VTS that satisfy the given properties. We improved the encoding of some of the conditions for VTSs. In the earlier implementation, reachability was encoded by enumeration of paths. In our current encoding, we have recursively defined the reachability condition such that that we avoid the exponential enumeration. We have also encoded $k$-connectivity in a similar fashion and avoided a blowup in the formula size. Furthermore, our encoding supports several variations of the conditions. We have implemented the encoding using Z3~\cite{z3} python API and searched for VTS satisfying various properties upto size 14-18 nodes as compare earlier tool that could only search graph upto size 8 nodes. Our tool is useful to cell biologists since VTSs of real cells have approximately ten compartments.  \\
